---
title: Stat 541 Experimental Design Project 1
author: "Ben DeVries, Eric Folsom, Ryan Kardoes, Alex Trauner"
format: pdf
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

```{r}
pre_dat <- data.frame(pot = rep(c("small", "large"), each = 3),
                      seconds = c(170, 162, 184, 112, 126, 119))
sig <- pre_dat |> group_by(pot) |> summarise(s = sd(seconds)) |> pull()
power.t.test(delta = 30, sd = sig[1], sig.level = 0.05, type = "two.sample",
             power = 0.90, alternative = "one.sided")
power.t.test(delta = 30, sd = sig[2], sig.level = 0.05, type = "two.sample",
             power = 0.90, alternative = "one.sided")
```

9-Step Checklist
1.	Define the objectives: Compare boiling times of water with and without salt to determine which boils faster. 
2.	Identify response and sources of variation. 
    a.	Treatment factors and levels: One treatment factor (salt) with 2 levels: unsalted, salted. 
    b.	Response: Time until boiling, in seconds. True difference is defined as 30 seconds.
    c.	Sources of variation: Pot material, stove used, burner used, time to preheat burner, amount of water used, amount of salt used, order of boiling (pot temperature).
    d.	Blocking factor: Pot + burner combination
3.	Choose a rule for assigning the experimental units to trt (design). 
    a.	We would choose a randomized block design, where salt treatment would be randomized so that each pot/burner combination received multiple repetitions of salted vs. unsalted in a random order.
4.	Specify measurements to be made, experimental procedure, and the anticipated difficulties. 
    a.	Measurements: Seconds until water is at a rolling boil (determined by one group member). 
    b.	Procedure: After randomizing salted vs. unsalted treatments to each pot and burner combination, begin the experiment. Burners should be preheated for 5 minutes prior to placing pots on burners. 2 cups of water should be added to pots. Unsalted pots can be boiled as is, salted pots should have 2 tablespoons of salt added prior to placing on the burner. Time should start when the pot is placed on the burner. Time should be recorded when the pot is at a rolling boil. In between repetitions, pots should be placed into a cool water bath to return them to neutral temperature before boiling the next pot of water.
    c.	Anticipated difficulties: Human subjectivity in identifying boiling point, ensuring pots and water are at same temp for different repetitions. 
5.	Pilot experiment: Our pilot experiment yielded the following results:
    a.	
6.	Specify the model: 
    a.	Seconds to boiling = constant + effect of salt + effect of burner and pot block + error 
    b.	Yijt = µ + αi + βj +εijt
7.	Outline the analysis: 
    a.	Compare differences between salted and unsalted water for time until boiling. Determine differences (confidence intervals) in boiling time for each treatment, and determine which treatment is best (fastest/least time until boiling). 
8.	Calculate number of observations and time/budget. 
    a.	N: Based on the our pilot study, we will need to complete
    b.	Time: Each pot will take 
    c.	Budget: Cost of salt + cost of electricity + cost of water + cost of human time 
9.	Review and revise. After the experiment, steps could be revised to plan a better future experiment. 


